---
title: "Hands-on Exercise 10"
author: "Goh Fang Yu"
date: "20 Jun 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **20 Information Dashboard Design: R methods**

## **20.1 Overview**

By the end of this hands-on exercise, you will be able to:

- create bullet chart by using ggplot2,

- create sparklines by using ggplot2 ,

- build industry standard dashboard by using R Shiny.

## **20.2 Getting started**

For the purpose of this hands-on exercise, the following R packages will be used.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse) 

```

- [tidyverse](https://www.tidyverse.org/) provides a collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. It is not a single package but a collection of modern R packages including but not limited to [readr](https://readr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [ggplot](https://ggplot2.tidyverse.org/), [tibble](https://tibble.tidyverse.org/), [stringr](https://stringr.tidyverse.org/), [forcats](https://forcats.tidyverse.org/) and [purrr](https://purrr.tidyverse.org/).

- [lubridate](https://r4va.netlify.app/chap31) provides functions to work with dates and times more efficiently.

- [ggthemes](https://r4va.netlify.app/chap31) is an extension of ggplot2. It provides additional [themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) beyond the basic themes of ggplot2.

- [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) provides some additional helper functions to assist in creating beautiful tables with [gt](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using the R programming language.

- [reactable](https://glin.github.io/reactable/index.html) provides functions to create interactive data tables for R, based on the [React Table](https://github.com/tanstack/table/tree/v7) library and made with [reactR](https://github.com/react-R/reactR).

- [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes.

## **20.3 Importing Microsoft Access database**

### **20.3.1 The data set**

For the purpose of this study, a personal database in Microsoft Access mdb format called *Coffee Chain* will be used.

### **20.3.2 Importing database into R**

In the code chunk below, `odbcConnectAccess()` of [**RODBC**](https://cran.r-project.org/web/packages/RODBC/index.html) package is used used to import a database query table into R.

```{r}
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con) 

```

Note: Before running the code chunk, you need to change the R system to 32bit version. This is because the `odbcConnectAccess()` is based on 32bit and not 64bit

### **20.3.3 Data Preparation**

The code chunk below is used to import *CoffeeChain.rds* into R.

```{r}
coffeechain <- read_rds("data/rds/CoffeeChain.rds")

```

Note: This step is optional if *coffeechain* is already available in R.

The code chunk below is used to aggregate Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()

```

### **20.3.4 Bullet chart in ggplot2**

The code chunk below is used to plot the bullet charts using ggplot2 functions.

```{r}
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()

```

## **20.4 Plotting sparklines using ggplot2**

In this section, you will learn how to plot sparklines by using ggplot2.

### **20.4.1 Preparing the data**

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)

```

The code chunk below is used to compute the minimum, maximum and end othe the month sales.

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))

```

The code chunk below is used to compute the 25 and 75 quantiles.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)

```

### **20.4.2 sparklines in ggplot2**

The code chunk used.

```{r}
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())

```

## **20.5 Static Information Dashboard Design: gt and gtExtras methods**

In this section, you will learn how to create static information dashboard by using [gt](https://gt.rstudio.com/) and [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) packages. Before getting started, it is highly recommended for you to visit the webpage of these two packages and review all the materials provided on the webpages at least once. You done not have to understand and remember everything provided but at least have an overview of the purposes and functions provided by them.

### **20.5.1 Plotting a simple bullet chart**

In this section, you will learn how to prepare a bullet chart report by using functions of gt and gtExtras packages.

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()

```

## **20.6 sparklines: gtExtras method**

Before we can prepare the sales report by product by using gtExtras functions, code chunk below will be used to prepare the data.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()

```

It is important to note that one of the requirement of gtExtras functions is that almost exclusively they require you to pass data.frame with list columns. In view of this, code chunk below will be used to convert the report data.frame into list columns.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

```

### **20.6.1 Plotting Coffechain Sales report**

```{r}
#| echo: true
#| results: 'hide'

report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)

```

### **20.6.2 Adding statistics**

First, calculate summary statistics by using the code chunk below.

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)

```

### **20.6.3 Combining the data.frame**

Next, use the code chunk below to add the statistics on the table.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

```

```{r}
sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )

```

```{r}
sales_data = left_join(sales, spark)

```

### **20.6.4 Plotting the updated data.table**

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline(
    column = "Monthly Sales",
    same_limit = FALSE,
    palette = c("white", "white", "purple", "green", "white")
  ) %>%
  tab_options(
    table.background.color = "white"
  )

```

### **20.6.5 Combining bullet chart and sparklines**

Similarly, we can combining the bullet chart and sparklines using the steps below.

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()  

```

```{r}
sales_data = sales_data %>%
  left_join(bullet)

```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline(
    column = "Monthly Sales",
    type = "points",
    same_limit = FALSE,
    palette = c("white", "white", "purple", "green", "white")
  ) %>%
  gt_plt_bullet(
    column = Actual,
    target = Target,
    width = 28,
    palette = c("lightblue", "black")
  ) %>%
  gt_theme_538()

```

## **20.7 Interactive Information Dashboard Design: reactable and reactablefmtr methods**

In this section, you will learn how to create interactive information dashboard by using [reactable](https://glin.github.io/reactable/index.html) and [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) packages. Before getting started, it is highly recommended for you to visit the webpage of these two packages and review all the materials provided on the webpages at least once. You done not have to understand and remember everything provided but at least have an overview of the purposes and functions provided by them.

In order to build an interactive sparklines, we need to install **dataui** R package by using the code chunk below.

```{r}
remotes::install_github("timelyportfolio/dataui")

```

Next, you all need to load the package onto R environment by using the code chunk below.

```{r}
library(dataui)

```

### **20.7.1 Plotting interactive sparklines**

Similar to gtExtras, to plot an interactive sparklines by using reactablefmtr package we need to prepare the list field by using the code chunk below.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))

```

Next, [react_sparkline](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) will be to plot the sparklines as shown below.

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)

```

### **20.7.2 Changing the pagesize**

By default the pagesize is 10. In the code chunk below, arguments defaultPageSize is used to change the default setting.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)

```

### **20.7.3 Adding points and labels**

In the code chunk below `highlight_points` argument is used to show the minimum and maximum values points and `label` argument is used to label first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)

```

### **20.7.4 Adding reference line**

In the code chunk below `statline` argument is used to show the mean line.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)

```

### **20.7.5 Adding bandline**

Instead adding reference line, bandline can be added by using the `bandline` argument.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)

```

### **20.7.6 Changing from sparkline to sparkbar**

Instead of displaying the values as sparklines, we can display them as sparkbars as shown below.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)

```

## **Exploration**

**1) Target Achievement Analysis using Bullet Charts**

Identify products that exceeded expectations and those that underperformed relative to their sales targets.

```{r}
target_perf <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(Product) %>%
  summarise(
    Target = sum(`Budget Sales`, na.rm = TRUE),
    Actual = sum(Sales, na.rm = TRUE),
    Achievement = Actual / Target,
    .groups = "drop"
  ) %>%
  arrange(desc(Achievement))

target_perf %>%
  gt() %>%
  fmt_percent(
    columns = Achievement,
    decimals = 1
  ) %>%
  gt_plt_bullet(
    column = Actual,
    target = Target,
    width = 35,
    palette = c("lightblue", "black")
  ) %>%
  gt_theme_538()

```

::: callout-tip
### Findings:

Overall sales performance was strong, with 10 of 13 products exceeding their targets. Earl Grey recorded the highest achievement (134%), while Decaf Irish Cream had the lowest (94.8%).
:::

**2) Seasonality Analysis using Enhanced Sparklines**

Identify seasonal trends, sales peaks, and low-demand periods that may influence inventory and marketing decisions.

```{r}
monthly_report <- coffeechain %>%
  mutate(
    Year = year(Date),
    Month = month(Date, label = TRUE, abbr = TRUE)
  ) %>%
  filter(Year == 2013) %>%
  group_by(Product, Month) %>%
  summarise(
    Sales = sum(Sales, na.rm = TRUE),
    .groups = "drop"
  )

seasonality_data <- monthly_report %>%
  group_by(Product) %>%
  summarise(
    Min = min(Sales),
    Max = max(Sales),
    Average = mean(Sales),
    `Monthly Sales` = list(Sales),
    .groups = "drop"
  )

seasonality_data %>%
  gt() %>%
  fmt_number(
    columns = c(Min, Max, Average),
    decimals = 0
  ) %>%
  gt_plt_sparkline(
    column = "Monthly Sales",
    type = "points",
    same_limit = FALSE,
    palette = c("white", "white", "purple", "green", "white")
  ) %>%
  gt_theme_538()

```

::: callout-tip
### Findings:

The sparklines show that most products exhibit recurring seasonal fluctuations, with sales generally rising toward the later months of the year. Colombian and Lemon recorded the highest average monthly sales, while Amaretto and Regular Espresso consistently generated the lowest sales volumes.
:::

**3) Executive Product Performance Dashboard**

Provide a consolidated view of product performance by combining sales trends, targets, and summary statistics in a single dashboard.

```{r}
spark <- monthly_report %>%
  group_by(Product) %>%
  summarise(
    `Monthly Sales` = list(Sales),
    .groups = "drop"
  )

summary_stats <- monthly_report %>%
  group_by(Product) %>%
  summarise(
    Min = min(Sales),
    Max = max(Sales),
    Average = mean(Sales),
    .groups = "drop"
  )

bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(Product) %>%
  summarise(
    Target = sum(`Budget Sales`, na.rm = TRUE),
    Actual = sum(Sales, na.rm = TRUE),
    Achievement = Actual / Target,
    .groups = "drop"
  )

dashboard_data <- summary_stats %>%
  left_join(spark, by = "Product") %>%
  left_join(bullet, by = "Product") %>%
  arrange(desc(Achievement))

dashboard_data %>%
  gt() %>%
  fmt_number(
    columns = c(Min, Max, Average, Actual, Target),
    decimals = 0
  ) %>%
  fmt_percent(
    columns = Achievement,
    decimals = 1
  ) %>%
  gt_plt_sparkline(
    column = "Monthly Sales",
    type = "points",
    same_limit = FALSE,
    palette = c("white", "white", "purple", "green", "white")
  ) %>%
  gt_plt_bullet(
    column = Actual,
    target = Target,
    width = 28,
    palette = c("lightblue", "black")
  ) %>%
  gt_theme_538()

```

::: callout-tip
### Findings:

Tea-based products (Earl Grey, Green Tea, Darjeeling, and Mint) were the strongest performers, exceeding 125% of their sales targets while maintaining consistent monthly sales trends. Conversely, Colombian recorded the highest average sales volume but underperformed against its target, highlighting a gap between sales volume and target achievement.
:::