---
title: "Hands-on Exercise 4 Part 4"
author: "Goh Fang Yu"
date: "2 May 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **8. Funnel Plots for Fair Comparisons**

## **8.1 Learning Outcome**

Funnel plot is a specially designed data visualisation for conducting unbiased comparison between outlets, stores or business entities.

By the end of this hands-on exercise, you will gain hands-on experience on:

- plotting funnel plots by using **funnelPlotR** package,

- plotting static funnel plot by using ggplot2 package, and

- plotting interactive funnel plot by using both **plotly R** and **ggplot2** packages.

## **8.2 Installing and Launching R Packages**

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)

```

## **8.3 Importing Data**

The code chunk below imports the data into R and save it into a tibble data frame object called *covid19*.

For this hands-on exercise, we are going to compare the cumulative COVID-19 cases and death by sub-district (i.e. kelurahan) as at 31st July 2021, DKI Jakarta.

```{r}
covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)

```

```{r}
knitr::kable(head(covid19, 10))

```

## **8.4 FunnelPlotR methods**

[**FunnelPlotR**](https://nhs-r-community.github.io/FunnelPlotR/) package uses ggplot to generate funnel plots. It requires a `numerator` (events of interest), `denominator` (population to be considered) and `group`. The key arguments selected for customisation are:

- `limit`: plot limits (95 or 99).

- `label_outliers`: to label outliers (true or false).

- `Poisson_limits`: to add Poisson limits to the plot.

- `OD_adjust`: to add overdispersed limits to the plot.

- `xrange` and `yrange`: to specify the range to display for axes, acts like a zoom function.

- Other aesthetic components such as graph title, axis labels etc.

### **8.4.1 FunnelPlotR methods: The basic plot**

The code chunk below plots a funnel plot.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)

```

Things to learn from the code chunk above.

- `group` in this function is different from the scatterplot. Here, it defines the level of the points to be plotted i.e. Sub-district, District or City. If Cityc is chosen, there are only six data points.

- By default, `data_type`argument is “SR”.

- `limit`: Plot limits, accepted values are: 95 or 99, corresponding to 95% or 99.8% quantiles of the distribution.

### **8.4.2 FunnelPlotR methods: Makeover 1**

The code chunk below plots a funnel plot.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",     #<<
  xrange = c(0, 6500),  #<<
  yrange = c(0, 0.05)   #<<
)

```

Things to learn from the code chunk above. + `data_type` argument is used to change from default “SR” to “PR” (i.e. proportions). + `xrange` and `yrange` are used to set the range of x-axis and y-axis

### **8.4.3 FunnelPlotR methods: Makeover 2**

The code chunk below plots a funnel plot.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",   
  xrange = c(0, 6500),  
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by \nCumulative Total Number of COVID-19 Positive Cases", #<<           
  x_label = "Cumulative COVID-19 Positive Cases", #<<
  y_label = "Cumulative Fatality Rate"  #<<
)

```

Things to learn from the code chunk above.

- `label = NA` argument is to removed the default label outliers feature.

- `title` argument is used to add plot title.

- `x_label` and `y_label` arguments are used to add/edit x-axis and y-axis titles.

## **8.5 Funnel Plot for Fair Visual Comparison: ggplot2 methods**

In this section, you will gain hands-on experience on building funnel plots step-by-step by using ggplot2. It aims to enhance you working experience of ggplot2 to customise speciallised data visualisation like funnel plot.

### **8.5.1 Computing the basic derived fields**

To plot the funnel plot from scratch, we need to derive cumulative death rate and standard error of cumulative death rate.

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate*(1-rate)) / (Positive))) %>%
  filter(rate > 0)

```

Next, the *fit.mean* is computed by using the code chunk below.

```{r}
fit.mean <- weighted.mean(df$rate, 1/df$rate.se^2)

```

### **8.5.2 Calculate lower and upper limits for 95% and 99.9% CI**

The code chunk below is used to compute the lower and upper limits for 95% confidence interval.

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
dfCI <- data.frame(number.ll95, number.ul95, number.ll999, 
                   number.ul999, number.seq, fit.mean)

```

### **8.5.3 Plotting a static funnel plot**

In the code chunk below, ggplot2 functions are used to plot a static funnel plot.

```{r}
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label=`Sub-district`), 
             alpha=0.4) +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll999), 
            size = 0.4, 
            colour = "grey40") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul999), 
            size = 0.4, 
            colour = "grey40") +
  geom_hline(data = dfCI, 
             aes(yintercept = fit.mean), 
             size = 0.4, 
             colour = "grey40") +
  coord_cartesian(ylim=c(0,0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") + 
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") + 
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") + 
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12),
        legend.position = c(0.91,0.85), 
        legend.title = element_text(size=7),
        legend.text = element_text(size=7),
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),
        legend.key.height = unit(0.3, "cm"))
p

```

### **8.5.4 Interactive Funnel Plot: plotly + ggplot2**

The funnel plot created using ggplot2 functions can be made interactive with `ggplotly()` of **plotly** r package.

```{r}
fp_ggplotly <- ggplotly(p,
  tooltip = c("label", 
              "x", 
              "y"))
fp_ggplotly

```

## **Exploration**

**1) Recovery Rate Funnel Plot**

```{r}
funnel_plot(
  .data = covid19,
  numerator = Recovered,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  x_label = "Total Positive Cases",
  y_label = "Recovery Rate",
  title = "COVID-19 Recovery Rate by Sub-district"
)

```

Observations:

- Most sub-districts are clustered closely around the overall recovery rate and fall within the control limits, suggesting relatively consistent recovery outcomes across the districts. However, several labelled sub-districts such as SENAYAN and KUNINGAN BARAT appear above the average recovery rate, while GANDARIA SELATAN falls below the lower control region, indicating comparatively weaker recovery performance.

- Although several sub-districts appear to deviate slightly from the overall recovery rate, most observations remain within the control limits, suggesting no strong evidence of extreme outliers. The relatively tight clustering of points also indicates broadly consistent healthcare recovery outcomes across the sub-districts.

**2) Fatality Rate Funnel Plot**

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.06),
  label = NA,
  title = "COVID-19 Fatality Rate by Sub-district",
  x_label = "Total Positive Cases",
  y_label = "Fatality Rate"
)

```

Observations:

The majority of sub-districts fall within the funnel boundaries, suggesting that observed differences in fatality rates are largely attributable to natural variation rather than extreme anomalies. However, a few sub-districts with relatively high fatality rates may indicate potential differences in healthcare accessibility, demographic vulnerability, or reporting practices that warrant further investigation.
